PanGraph: scalable bacterial pan-genome graph construction \\ Supplementary Information

### 2023-03-08 reviewer_response.tex

\reviewer{R2.6}{Line 183 - I really do not understand what is going on with your alpha, beta being set to zero. Are you saying if you stop penalising merges for having too many SNPs, then you will be able to merge more, and this will stop you from havign a 10\% divergence limit? It will also mean you make more false merges, right?}
% \Marco{setting $\beta$ to zero will indeed remove the 10\% divergence limit set by the pseudo-energy. There is another limit that depends on the aligner sensitivity (20 to 30\%, depending on the aligner). This limit remains, and in this sense no ``false merges'' are performed? Not sure I interpret this correctly.}\\
% \Liam{I think your interpretation is right. Important to clarify that there are those two limits and that $\beta$ only controls the divergence limit \textit{within} what one can actually align.}
\response{There are two factors limiting the divergence of mergers that are performed. The first factor is the aligner used. Minimap2 is in general able to find matches up to roughly 20\% sequence divergence, while mmseqs2 can reach divergence up to roughly 30\%. Only homology matches of up to this divergence will be suggested by the aligners.
      On top of this, only matches with negative pseudo-energy are accepted as mergers.
      The parameter $\beta$ thus allows to explicitly set a maximal divergence ($1/\beta$) allowed in mergers, rather than relying on the implicit cut-off that comes from the sensitivity of the aligner. \\
      This is now made more explicit in SI sect.~{\SIalgo}C.\\
      When setting $\beta = 0$ the maximal divergence is determined by the aligner sensitivity. In this sense more merges are allowed, but all mergers will have some implicit maximal divergence threshold (20\% or 30\%, depending on the aligner).}

\reviewer{R2.7}{Why does the L50 peak go up for null-scores for mmseqs for Mtb, Pm, Kpn and Ec (Figure 4E, purple with lines) but down for helicobacter?}
% \Marco{the effect of merging more blocks on L50 is not trivial. It depends on whether the merging occurs between long or short blocks, and how much fragmentation this causes. HP is a special case in that its diverge is on the threshold of what minimap2 can handle. Using mmseqs2 results in a pangenome that is half the size compared to standard parameters (fig. 4G). This is an impressive amount of additional compression. At the same time the fraction of core pangenome more than doubles (4H). The total number of blocks decreases to roughly half (4D) while N50 increases (4F). For me this points to the fact that in the case of HP merging will split long blocks and create more fragmentation. For the reviewer I could produce the block frequency and lenght distribution for the two cases and comparing them would give an answer. I could do the same for EC or KP for comparison.}\\
% \Liam{makes sense.}
\response{This is an interesting observation. L50 measures the minimum number of pancontigs needed to cover at least 50\% of the pangenome graph sequence. This number is closely related to the total number of pancontigs. In general the use of null energy parameters result in more fragmented graphs, with a higher number of pancontigs (cf. fig.~{\SIfigBenchmark}D in SI) so L50 increases. However, the reviewer is correct that \textit{Helicobacter} is an exception and shows the opposite trend: the use of null energy parameters results in less fragmentation of the graph (both in the number of pancontigs and for the value of L50).\\
      Having investigated, we believe this might be explained by that fact that the divergence of some parts of the \textit{Helicobacter} core genome in this dataset is close to $1/\beta$ for $\beta=10$ used in the figure.
      Hence, parts of the core genome where the divergence reaches $>1/\beta$ might be split into different pancontigs. Removing the threshold imposed by $\beta$ by setting $\beta=0$ results in a longer consecutive merger and therefore a less fragmented graph with fewer overall pancontigs.\\
      For a visual verification of this interpretation we display in fig.~\ref{fig:kp-vs-hp} how the distributions of pancontig length and frequency change with the use of different alignment kernels and pseudo-energy parameters: minimap (asm10) with default parameters vs. mmseqs with null pseudo-energy. We compare the changes observed on the Hp dataset to the ones in Kp. In the Kp dataset the null pseudo-energies result in shorter pancontigs (lower-right panel) and a small increase in core genome fraction (lower-left panel). For Hp, the effect on pancontigs is quite different. More sensitive alignment with null pseudo-energies means small and intermediate pancontigs are lost, but there are slightly more long ones (top-right panel). It also results in a loss of low-frequency pancontigs (top-left panel) indicating that these had an homologous counterpart that with default parameters and minimap could not be merged due to high-divergence. Using a more sensitive aligner with null pseudo-energies means the core pangenome fraction increases considerably.  
      % TODO: add mention in SI
      % QUESTION: this is more or less explicit in SI? Should I expand more?
      % Liam: I think there are slightly more long pancontigs with mmseqs/null in Hp from looking at figure so seems worth saying
}

@@

\begin{figure}[htb]
      \includegraphics[width=.7\textwidth]{figs_response/hp_vs_kp.pdf}
      \caption{frequency and length distribution (respectively left and right panels) of pancontigs for the \textit{Helicobacter pylori} and \textit{Klebsiella pneumoniae} datasets (top and bottom). For each panel we report the results for pangenome graphs built using minimap2 aligner with standard parameter values and asm20 sensitivity (blue) and with mmseqs2 and null pseudo-energy parameters (orange). Distributions are weighed by pancontig length, so that the area under the curve represents the total pangenome graph size.}
      \label{fig:kp-vs-hp}
\end{figure}

\reviewer{R2.8}{Just to make sure i understand - is it possible for sequence to be ``left out'' in between pancontigs?}
% \Marco{no, every part of the genome must be contained in a pancontig. It can be left out from the consensus if it is considered an insertion.}\\
% \Liam{if they possibly misunderstood, suggests that some statement that all sequence ends up in a pancontig should be added to manuscript. I suspect they are thinking of case of short ($<$100bp) stretches `in between' pancontigs, which (if I understand correctly) will be incorporated as insertions at the end of the consensus alignment of the adjacent pancontig.}
\response{No, every part of the sequence of each genome must be included in some pancontig. It might not necessarily appear in the pancontig consensus sequence if it is present in the alignment as a short ($< \Lthr$) insertion, but it will nonetheless be present in the alignment. In the particular case in which merging two pancontigs would leave overhangs shorter than $\Lthr$, these are included in the alignment as flanking insertions. This is now explained in SI section {\SIgraph}, and mentioned in the main text in section IB (line 103).
      In fact, we used the fact that genomes can be reconstructed exactly from the pangraph to test and debug the tool. }

@@

\reviewer{R2.9}{About data accessibility
      \begin{enumerate}
            \item i don't see how to get the list of accessions for section IIB? Saying they were used in the Ding et al paper is not quite enough for me to be sure which ones you used.
            \item for the graph marginalisation section, could the accessions of the strains used be in a supplementary file as well as on github? Github repos are great, but they can be deleted.
      \end{enumerate}}
% \Marco{on the repo we have \href{https://github.com/neherlab/pangraph/blob/master/script/config/accnums.json}{this file} for the list of all accession numbers we use, and \href{https://github.com/neherlab/pangraph/blob/master/script/config/projection_strains.json}{this file} for the marginalization pairs. I could create some excel spreadsheet with accession numbers that we could add to the submission?}\\
% \Liam{for sure that's all they want (most biologists don't like json!)}
\response{We now included in the submission two csv files. \texttt{accession\_numbers.csv} contains the list of accession numbers of all isolates in our dataset for all species considered. \texttt{marginalize\_pairs.csv} contains instead a list of all the pairs used in the graph marginalization section.
      % TODO: archive repository state.
}

\subsubsection*{Minor Issues}

\reviewer{R2.1}{line 13 - it would be good to give a reference for the statement that even closely related genomes can differ widely in gene content.
      I don't know what the first paper to show this was, the earliest i am aware of is from Eduardo Rocha's team, which annpoyingly i can't locate right now, but Figure 2 here shows it \url{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1008866} \cite{touchon2020phylogenetic} - you can be very close in patristic distance and have very different gene repertoire.
      Feel free to replace with a better reference, this si just meant to be helpful.}
% \Marco{I think this is a good reference, fig. 2 is familiar to us. And also fig.2 in \cite{doolittle2009origin} which is a bit older. But not sure which other Rocha paper they are referring to.}\\
% \Liam{I think they mean Figure 5 of \cite{touchon2009organised} (\href{https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000344}{here}). Citing Touchon 2009 and 2020 and the Doolittle paper seems fine to me.}
\response{We thank the reviewer for the suggestion. We now cite the suggested article, along with \cite{touchon2009organised,doolittle2009origin}.}

\reviewer{R2.2}{Why on earth is the ``get an MSA of a block'' option in pangraph called ``polish''?}
% \Marco{i don't think this comes from the paper (I used grep on the tex file and we never mention polish). I think it comes from the documentation but not sure where. We explain polish \href{https://neherlab.github.io/pangraph/tutorials/tutorial_3/\#Polishing-the-alignments}{in this part} of the tutorial and I don't think it's ambiguous. If in the paper we explain how we propagate the consensus sequence and related problems we will also mention the polish command in the same part and make explicit that it can be used to polish alignments, this should solve the ambiguity.}\\
% \Liam{this sounds good. I think they misunderstood that polish is not just `extract the MSA' but actually produces a pangraph with the same structure but redone alignments. I think what is in the documentation there is great and could almost go straight in the paper, something like: `At each stage pancontigs are compared and aligned using their consensus sequences. This shortens considerably the time needed, but might introduce minor inconsistencies and artifacts in the alignments. These can be removed at the end of the process by reconstructing the sequences in the block and performing a multiple sequence alignment (see Polish command in documentation).')}
\response{In our method pancontig mergers are performed using only the consensus sequence as a guide. This makes the merging procedure very fast, but can introduce minor inconsistencies and artifacts with insertions in the alignment as insertions are not aligned to each other in the procedure.
      To remove these inconsistencies, after building the graph the user can optionally decide to ``polish'' pancontig alignments with the \texttt{polish} command. This command takes a graph as input and produces a graph with the same topology, but with improved alignments. In practice for each block it collects all of the sequences and uses MAFFT \cite{katoh2002mafft} to perform a multiple sequence alignment, obtaining an improved consensus and more consistent insertion pattern.\\
      This command was not mentioned in the paper. This has now been amended, and we briefly described in in section IB in the main text and in SI sect.~{\SIalgo}B.}

\reviewer{R2.3}{line 48: ``Pancontigs are connected by an edge if they are syntenic on at least one input sequence;" - i think this is a bit of an abuse of modern language. The word syntenic is generally used to mean two genes/whatever have order conserved on multiple chromosomes (although some googling tells me that in classical genetics it meant things were on the same chromosome). Anyway, my understanding is pancontigs are connected by an edge if one of the input genomes is spelled out by going from one to the other. So it's about adjacency rather than synteny.}
% \Marco{I agree, we could use ``if they are ajacent/contiguous on at least one input sequence''?}\\
% \Liam{adjacent sounds good.}
\response{We agree with the reviewer's observation. We have changed ``syntenic on'' to ``adjacent in''.}

\reviewer{R2.4}{Line 131: was N really=100? on the plot in figure 2, you have n.isolates going up to 1000}
% \Marco{$N=100$ was the standard value for the rest of the simulations (e.g. later for accuracy). I did a tentative correction: removed from the main the specification on population size and mentioned to check the SI for details on standard parameter values.}
\response{The value $N=100$ is the standard value used in the rest of the simulation, but as the reviewer pointed out this value is varied for the results showed in fig.~2 and the statement was confusing. We modified it removing the value of $N$ and referring to SI section {\SIsimBench} for the standard value of simulation parameters.}

@@

\subsection*{Reviewer 3}

\reviewer{R3}{The article ``PanGraph: scalable bacterial pan-genome graph construction'' reports a new method for constructing a Pan-Genome Graph (PGG) in an efficient manner with an open and extractable grah structure. Unfortunately, the authors appear to either be unaware of or downplay the existence of previous methods for the same purpose. The authors need to survey the existing literature and compare and contrast with their method to show the utility of their approach.\\
      For instance the authors state: ``Traditional computational approaches towards whole-genome analysis either scale poorly with the number of genomes, or treat genomes as dissociated ``bags of genes'', and thus are not suited for this new era.''. There are many pan-genome analysis tools which treat the pan-genome as a graph and generate a PGG. There are two primary differences in these existing tools: those which generate a graph at the nucleotide level with nodes representing basepairs (sometimes nonbranching runs of nodes are collapsed) and those where nodes represent larger features such as genes.\\
      PanGraph appears to wed these approaches to some extent in that the alignments are done at the nucleotide level and the features are extracted as a characteristic of the alignments.}

\response{
      We apologize if our previous discussion of the literature was incomplete. It was not our intention to downplay prior work.
      We thank the reviewer for highlighting additional important work and aspects that we have now added to better contextualise our work.
      We have rewritten the introduction to reflect previous work in this field and added  to the discussion to further highlight previous and alternative approaches. We hope that these edits address the concerns of the reviewer.
      % Liam: I think good to say in more detail what has been added here. 
}

\reviewer{R3}{An obvious lack in PanGraph is an attempt to map existing annotations back onto the generated PGG which is a necessity for usefulness to biologists. The mechanism for this is obvious but providing a useful user interface might be tedious.}
% \Liam{I think perhaps producing a helper python script that converts a genbank annotation file for a genome within a pangraph to an output gff/csv with the block locations and coordinates would please this reviewer.}\\
% \Marco{Yeah good idea. Maybe something that just map each annotation into a block, without solving the problem of harmonizing annotations which might be some more work...}\\
\response{We thank the reviewer for this important suggestion and appreciate the need to include a way to map annotations back onto the pangenome graph. We created a helper script that allows a mapping of pancontig information onto a set of annotations in a \texttt{gff} file. In default usage the pancontig location(s) of each feature gets added as a new `attribute', allowing a quick connection between the generated pangenome graph and existing annotations. The script and instructions on its usage can be found at \url{https://github.com/liampshaw/pangraph\_annotations}. In the future, we hope to extend this further for visualization purposes.
      % TODO: mention this in the text.
      % Liam: not sure if we need last sentence
}

\reviewer{R3}{
      This statement: ``While easy to conceptualize, the construction of pangenome graphs has proven computationally challenging. Colored generalizations of the de Bruijn graph-based assemblers have been successively used to build graphs from large sequence sets, although the underlying efficiency derives from a fixed k-mer size which prevents modelling long-range homology (Iqbal et al., 2012; Muggli et al., 2017).'' appears to be incomplete at best. De Bruijn graph-based assemblers routinely collapse nonbranching portions of the graph into longer sequences than the k-mer size used for construction.\\
      One early approach that does not explicitly generate a pan-genome graph but deals with some of the rearrangement issues cited by the authors is:
      \begin{itemize}
            \item \cite{angiuoli2011improving}
            \item \cite{angiuoli2011mugsy}
      \end{itemize}
      While the underlying approach is quite different, a very similar PGG to PanGraph is produced by a different method with similar nodes represented by multiple sequence alignments and edges with the same meaning:
      \begin{itemize}
            \item \cite{sutton2021pan}
            \item \cite{chan2015novel}
      \end{itemize}}
\response{
      We apologize and we modified the sentence in question. We also extended the discussion of de Bruijn graphs methods in the introduction.\\
      We thank the reviewer for pointing out these important contributions that were not previously discussed in the paper.
}
% \Marco{from what I understand the structure in de Bruijn graphs is very fine-grained. Every SNP can generate a branching event. While in our case we have this intermediate corse-graining where homologous parts are collapsed together and the remaining structure of the graph represents large-scale structural variation in the genomes. I don't think this is solved by collapsing nonbranching portions in dB graphs (but I might not know enough about this...)}\\

\reviewer{R3}{I believe this statement: ``It is not uncommon that one pancontig has homology with multiple other pancontigs and the iterative algorithm has to make choice which potential mergers are performed and in which order.'' is the essential step in the process and perhaps deserves more focus. It captures two important issues: aligning orthologs not paralogs and keeping the fracturing of the PGG nodes to a minimum. The pseudo-energy captures the alignment length (longer is better for orthologs versus paralogs) but does not appear to capture the surrounding context of the iterative PGGs. For initial alignments of single genomes to single genomes this doesn't matter for the most part because there is a single pancontig for each each genome and PGG context is mostly captured in the alignment. I am not so sure this is true as the PGG nodes fracture and become much smaller. When choosing between competing mergers of paralogs/orthologs the context of the PGG and other possible mergers is important if the pancontigs have been reduced so that they contain only repetitive sequence without unique flaning sequence. I think this is a very important point as distinguishing orthologs from paralogs is essential for correct PGG construction and the authors need to explicitly address it. This could be measured as part of the simulated data where identical paralog sequences of various lengths and copy number could be modeled.}
% \Liam{I am not sure I agree that constructing a pangenome graph means one must distinguish orthologs from paralogs. The structure is still useful even if this isn't done. Maybe there is a way of doing something of what they propose with simulations, but I suspect this isn't necessary for the editor to accept the paper\ldots}\\
% \Marco{I agree, I think it's a different design choice. We could also mention that the marginalize command has an option to project out paralogs, but should be a post-processing step rather than something that we want to do while building the graph. One can a posteriori decide whether one wants to split paralogs using structural information (flanking blocks) or one can even prevent the merging using diversity and the $\beta$ parameter.}\\
\response{
This is an important point that needed more discussion in the paper than we originally provided. To be clear: \textit{PanGraph} in its current state does not explicitly distinguish between orthologs and paralogs when constructing the graph. 
This is a deliberate design choice. % Liam: don't understand this part of sentence so deleted 'as better explained in the answer to the last question.'. I think it's better to say that it's deliberate than to equivocate
The possibility of grouping duplicated sequences in the same pancontig will also group information on their context, and the fragmentation introduced in the graph would reflect some real (potentially ancient) structural change in the genome.\\
% Liam: not sure I quite understand above sentence...

### 2023-03-08 reviewer_response.tex

However, as the reviewer points out, in the case of distantly related paralogs such a grouping might be undesirable if it introduces `complications' in the graph. We agree from experience that these can sometimes make understanding the pangenome more complicated. Currently \textit{PanGraph} supports two different ways of controlling paralog splits.\\
The first one is based on sequence divergence. By tuning the value of the energy parameter $\beta$ the user can set a divergence limit. This would prevent the merging of paralogs that have sequence divergence higher than $1/\beta$.\\
The second method relies instead on information on genomic context. The \texttt{marginalize} command possesses the optional \texttt{--reduce-paralog} flag. When set, this flag will result in the removal of artificial splits caused by duplicated pancontigs that always occurr in the same non-duplicated context. These duplicated pancontigs are then split and absorbed in their context.\\

In the future we plan on providing a dedicated command to perform a more robust form of paralog splitting, which combines both sequence divergence information and context information. This command would build a phylogenetic tree from the alignment of each duplicated pancontig to detect exceptionally long branches, and check whether the split correlates with a split in the duplication context. This functionality is not however not yet implemented in the current version of \textit{PanGraph} and we see it as substantial extra work.\\

We appreciate the reviewer's perspective and have added SI sect.~{\SIalgo}D to include discussion of these considerations.

@@

\reviewer{R3}{How are pancontig consensus sequences determined and updated? I could not find any details about this. This is not a trivial operation. My reading of the article seems to imply that a complete multiple sequence alignment is not done for each merger. Does the consensus sequence degenerate iteratively with mergers? For the simulated data how accurate are the final pancontig consensus sequences?}
% \Liam{expanding discussion of doing an MSA at end as for R2's comments would partially address. Quantifying how the consensus sequence degenerates iteratively is an interesting point that seems possible?}\\
% \Marco{I agree, this could be addressed by adding a more detailed algorithm explanation of the merging algorithm in the SI. The short answer is that at each merging the full alignment is constructed using the CIGAR of the alignment of the consensus of the two blocks. Then consensus of the full alignment is re-evaluated. This does not solve the multiple alignment problem during constructions and there can be degeneration on where indels are mapped, but the final graph can be eventually polished. I don't think we ever quantified how much the consensus sequence degenerates during the iterations. Maybe one could devise a proper test set and see how it degenerates over the iterations.}\\
\response{This point was raised by other reviewers, and we appreciate the need to make this part of the alorithm clearer. To this end we added section {\SIalgo} in the SI (see ``graph merging''), with a detailed explanation of how the merging procedure works.\\
      In essence, during graph merging alignment is only performed between the consensus sequences of the merged pancontigs. We avoid performing a multiple sequence alignment during merging to maintain the algorithm scalable. However we do update the consensus of the pancontig at each merge using the majority rule on the merged alignment.\\
      The idea behind this approach is that the consensus sequence is needed to find large-scale homology, and resolving the fine-grained structure of the alignment is not needed at this step and would require an exccessive amount of resources. However after the graph has been built alignments can be refined with \textit{PanGraph}'s \texttt{polish} command. This command performs a multiple sequence alignment on all the sequences in each pancontig and updates the consensus sequence accordingly. This separates efficiently the problem of resolving the large-scale and nucleotide-level diversity between sequences.\\
      We did not quantify explicitly the divergence of consensus sequences during graph building, but we observed that, even without ``polishing'', alignments of up to hundreds of sequences look consistent.
      % TODO: quantify this in simulations if there is time?
      % Liam: Think the last sentence is potentially risky - I don't get what `consistent' means? My rewrite (not confident in this): In our experience, the alignments appear reliable throughout this process, although we have not explicitly quantified it from our simulations. However, for cases where the alignments are of interest we recommend running `polish' to perform MSA for pancontig consensus sequences (as in the documentation) to ensure that the final alignments are as accurate as possible.  
}

\reviewer{R3}{The statement: "Upon completion, transitive 108 edges within the graph, i.e., edges along which all path are co-linear, are removed by merging adjacent pancontigs." needs clarification. Why would these pancontigs have been split in the first place? Is there any necessary special issues with producing the merged multiple sequence alignment?}
% \Liam{don't understand their last sentence. Maybe should quantify how common the transitive edges typically are?}\\
% \Marco{I can think of one edge case which is due to periodic boundary conditions. Initially each circular genomes is saved as a single long block. If during alignment this unique blocks is split somewhere in the middle, the resulting edge blocks might be always colinear, since they are actually adjacent parts of the sequence that were artificially split to linearize the circular genome. One wants to collaps these two parts in a single block. This should happen often at the beginning of the merging of circular genomes. But might be unnecessary detail when explaining the algorithm the first time, we might want to move it to the SI section where we explain merging in more detail?}\\
\response{Transitive pancontigs are a small but problematic artifact of periodic boundary conditions when merging circular genomes, such as bacterial chromosomes. At the beginning of the procedure the circular genome is saved as a single linear pancontig. This introduces an artificial cut. When during merging this unique pancontig is further split, the flanking pancontigs could be crossing the artificial cut in a co-linear way on all paths. This step of the procedure resolves the artificial cut by merging these pancontigs together.\\
      We think that the mention of this step of the algorithm in this part of the paper might be a bit confusing and needs more clarification. We moved and expanded the statement from section IB of the main text to SI section {\SIalgo} (point 5 of the merging algorithm explanation).}

@@

% To amend this we added the size of soft-core genome for each species in table~1 in the main text, and added part of this discussion in SI sect.~{\SIdataBenchmark}A.
% TODO: core genome for simulated data if there is time?

%     "all core": 778,
%     "all softcore": 2081,
%     "all genes": 23128,
%     "all nondupl": 18798,
%     "drop core": 822,
%     "drop softcore": 2069,
%     "drop genes": 23054,
%     "drop nondupl": 18749

\reviewer{R3}{The authors might want to consider some form of post processing of the PGG to collapse what have been called bubbles in the graph if alignment of the pancontigs on different edges of the bubble warrant it.}
% \Marco{if I understand correctly this is hinting at some form of post-processing step in which artifactual small blocks - generated at the edges of long block merges - could be absorbed in the flanking block alignments and removed?}\\
\response{
      We appreciate the importance of the point raised by the reviewer.
      We agree with the reviewer that this post-processing smoothing of the graph produced by of \textit{PanGraph} would be a valuable addition, and we hope to such further post-processing refinements in subsequent versions. 
      This sort of refinement would allow alignments of adjacent edges of pancontigs to be harmonized, potentially also removing very small pancontigs when these could be appropriately absorbed in the alignment of larger ones. We hope to continue to expand the downstream post-processing of \textit{PanGraph} data structures in future work. 
      % Liam: tried to make it sound more explicitly 'future' and 'further'
}

@@

      % Other than being introduced by the diversity of the dataset, fragmentation could also be introduced artificially in the merging procedure. We tested for excessive fragmentation in simulated data using partition entropy, comparing the real underlying structural partition to the one reconstructed by pangraph (not shown in the paper). We found that fragmentation was present only when the diversity of the sample was too high for the sensitivity of the aligner.\\
      Irrespective of its origin, a possible way to detect excessive fragmentation and check for consistency in the graph structure is to visualize the distribution of pancontig size in the graph (see for instance fig.~\ref{fig:kp-vs-hp} lower right panel). Excessive fragmentation in the graph would be detectable as most of the sequence being condensed in pancontigs having size close to $\Lthr$.\\
      % TODO: show distributions for pangenome graphs with 500/1k isolates? Mention times?
      % TODO: add this discussion in the text?
      % TODO: mention the partition entropy? This could be an argument against artificial fragmentation.
      % TODO: maybe add in the SI the block frequency distribution for different species/kernels?
}

\reviewer{R3}{An issue I have struggled with in my own work is how to represent an orthologous gene cluster (node/vertex) within the PGG when the gene sequence in some genomes has an insertion event such as an IS element within the gene. I assume for PanGraph that this would necessitate splitting the node into multiple pancontigs. How is this sustainable for species with large numbers of active IS elements? This could also be investigated for PanGrah using suitable simulated data.}
% \Liam{I think nothing to say here apart from `yes, active IS elements split pancontigs up' (and can themselves be split up due to similarity with other IS\ldots). For the future one might want to have pangenome methods where you can explicitly pass a list of predefined sequences that should not be split. But I think this is a general problem and not something to investigate with simulations or fix here.}\\
% \Marco{agree. I think this is a structural property that we want to capture, the fact that these pancontigs are split.}
\response{
      We agree with the reviewer's observation that events such as the presence of an IS in an orthologous gene cluster would split the cluster in multiple pancontigs. For datasets in which isolates have a high amount of IS (longer than $\Lthr$), this would indeed result in multiple splits in the graph structure.\\
      This is in part an explicit design choice in \textit{PanGraph}. We developed this tool for the study of the evolution of bacterial pangenome, and we wanted this feature of the pangenome - namely the fragmentation due to presence of the same IS in multiple contexts - to be encoded in the graph. By condensing all occurrences of such an IS in a single pancontig we also condense the information on all of the different contexts in which the said IS appears (at the cost of fragmenting the graph). We agree that analysing these kinds of regions is challenging. In ongoing work we have been investigating the similarity of IS contexts to obtain insights into the evolution of IS-rich regions. This is especially important given the fact that the sequences of the IS themselves have very little phylogenetic information.\\
      However, we agree with the reviewer that this representation is not always desirable. One option to deal with it would be to allow the user to specify a set of `protected sequences' that could be used to filter out mergers involving e.g. exclusively ISs. Though this is in contrast to the agnostic approach of \textit{PanGraph} to sequence annotation, we are considering adding something like this in the future. However, this will require further research and potentially database curation at the level of species, so is not included in the current version.
}
